Quantifying and Modeling Global Water Accessibility including Water Typology

### 2021-01-29 docs/manuscript/manuscript.tex

\section{Contribution}

@@

Next, there were many missing values from the data. Therefore, we had to do a simple missing value analysis. We determined to avoid using water, transportation, or wealth variables which had greater than 50\% missing values. Ultimately, simplified data frame was 76 countries by 28 variables. After, two different tables are generated where one focused on the water accessibility variables and a explanatory variables. We selected 11 water accessibility variables based on the sources available such as location, time, and methodology \citep{price2019difference}. The water accessibility variable was organized
in the following table to allow for quick visualization for \autoref{tab:summary-wa}. The other variables were used as an explanatory variables for determining the water accessibility predictors in \autoref{tab:summary-var}. 
\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} p{2in}}\toprule
    \bf Variable & \bf Description & \bf Definition \\\midrule
    \texttt{ptap} &  Public tap/standpipe
 & 
 \\\midrule
 \texttt{bore} & Tube well/borehole & 
\\\midrule
\texttt{pspr} & Protected spring
& 
\\\midrule
\texttt{uspr} & Unprotected spring
& 
\\\midrule
\texttt{pwel} & Protected well
& 
\\\midrule
\texttt{uwel} & Unprotected well water
&
\\\midrule
\texttt{surw} & Surface water
& 
\\\midrule
\texttt{tles} & Water 30 minutes or less away round trip
&
\\\midrule 
\texttt{tmor} & Water more than 30 minutes away round trip & 
\\\midrule
\texttt{rain} & Rainwater
& 
\\\midrule
\texttt{phom} & Piped into dwelling

@@

\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} p{2in}}\toprule
    \bf Variable & \bf Description & \bf Definition \\\midrule
    \texttt{imws} & Improved water source
 & 
 \\\midrule
  \texttt{pipy} & Water piped into yard/plot
&
\\\midrule
 \texttt{truc} & Tanker truck & 
\\\midrule
 \texttt{ctan} & Cart with tank & 
\\\midrule
 \texttt{bott} & Bottled water & Drinking water is bottled water/demi john.
\\\midrule
 \texttt{uiws} & Unimproved water source & 
\\\midrule
 \texttt{othw} & Other water source & 
\\\midrule
 \texttt{iwsp} & Improved water source on the premises & The de jure population living in households with an improved water source on the premises.
\\\midrule
 \texttt{basw} & Basic water service
 & Defined as an improved water source with either water on the premises or round-trip collection time is 30 minutes or less.
\\\midrule
 \texttt{liws} & Limited water service
 & Limited water service, defined as an improved water source with round-trip collection time greater than 30 minutes.
\\\midrule
 \texttt{prem} & Water on the premises
 & 
\\\midrule
 \texttt{bicy} & Bicycle
 &
\\\midrule
 \texttt{acar} & Animal drawn cart
 & 
\\\midrule
 \texttt{mcyc} & Motorcycle
 & 
\\\midrule
 \texttt{pcar} & Private car
 & 
\\\midrule
 \texttt{boat} & Boat with a motor
 & 
\\\midrule
 \texttt{wigc} & Wealth index Gini coefficient
 & The Gini coefficient indicates the level of concentration of wealth in the country.
  \\\bottomrule
  \end{tabular}
  \caption{Summary of explanatory variables (National Household usage Averages measured as percentage for Main Drinking Source)}
  \label{tab:summary-var}
\end{table}
 \subsection*{Summary of variables}
\hl{Here, you show and discuss the data distributions (e.g. histograms, correlation plots, etc)}

@@

\subsection{Preamble}
We analyzed data from the Demographic Health Survey (DHS) and the Malaria Indicator Survey (MIS) (1990-2020) to highlight water accessibility predictors in 78 countries throughout the world.

@@

\subsection{Model results}
First, we developed a correlation chart to quickly analyze if there is any great correlation between all the 28 numerical variables that may pertain to water accessibility. We recognized the strong and weak relationship by the size and color of the correlation plot is shown in \autoref{fig:cor-basic}, which indicates households with limited water service has the strongest and positive relationship with Households possessing a bicycle. Meanwhile, households using an improved water source has the strongest and negative relationship with households using an unimproved water source. This initial finding results indicated that the initial variables are logical and can give us insight into further understanding and the relationship between these variables.

@@

\hl{Show the typology/cluster results and discuss.}
NBClust statistic analysis developed and determined four distinct groups. The first cluster group is composed of countries near the equator or in the Sub-Sahara Regions. The analysis of the second group represents countries considered developed or recently industrialized. The third cluster group represents most of the countries in central Africa and may have similar factors as the first cluster group. The fourth cluster group represents most of the countries in the middle east or predominately non-African countries. There are great internally heterogeneous clusters for the 76 distinct countries. The cluster analysis' dendrogram is shown in \autoref{fig:dendrogram}, and the corresponding NBClust curve yielding the optimal four clusters is shown in \autoref{fig:optimal-clusters}, \autoref{fig:optimal}. \autoref{tab:water-cluster} shows the resulting four water accessibility country clusters. The cluster's water accessibility variables are summarized in \autoref{tab:summary-wa}. \autoref{fig:cluster-map} shows the cluster country membership of the water accessibility on a world map.  Through the optimal analysis of the clusters, which indicated the superior grouping of four cuts along with the relevant indicators.

>>>>

\section{Background}

@@

Next, there were many missing values from the data. Therefore, we had to do a simple missing value analysis. We determined to avoid using water, transportation, or wealth variables which had greater than 50\% missing values. Ultimately, simplified data frame was 76 countries by 28 variables. After, two different tables are generated where one focused on the water accessibility variables and a explanatory variables. We selected 11 water accessibility variables based on the sources available such as location, time, and methodology \citep{price2019difference}. The water accessibility variable was organized
in the following table to allow for quick visualization for \autoref{tab:summary-wa}. The other variables were used as an explanatory variables for determining the water accessibility predictors in \autoref{tab:summary-var}. 
\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} p{2in}}\toprule
    \bf Variable & \bf Description & \bf Notes \\\midrule
    \texttt{ptap} &  Public tap/standpipe
 & 
 \\\midrule
 \texttt{bore} & Tube well/borehole & 
\\\midrule
\texttt{pspr} & Protected spring
& 
\\\midrule
\texttt{uspr} & Unprotected spring
& 
\\\midrule
\texttt{pwel} & Protected well
& 
\\\midrule
\texttt{uwel} & Unprotected well water
&
\\\midrule
\texttt{surw} & Surface water
& 
\\\midrule
\texttt{tles} & Water 30 minutes or less away round trip
&
\\\midrule 
\texttt{tmor} & Water more than 30 minutes away round trip & 
\\\midrule
\texttt{rain} & Rainwater
& 
\\\midrule
\texttt{phom} & Piped into dwelling

@@

\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} p{2in}}\toprule
    \bf Variable & \bf Description & \bf Notes \\\midrule
    \texttt{imws} & Improved water source
 & 
 \\\midrule
  \texttt{pipy} & Water piped into yard/plot
&
\\\midrule
 \texttt{truc} & Tanker truck & 
\\\midrule
 \texttt{ctan} & Cart with tank & 
\\\midrule
 \texttt{bott} & Bottled water & Drinking water is bottled water/demi john.
\\\midrule
 \texttt{uiws} & Unimproved water source & 
\\\midrule
 \texttt{othw} & Other water source & 
\\\midrule
 \texttt{iwsp} & Improved water source on the premises & The de jure population living in households with an improved water source on the premises.
\\\midrule
 \texttt{basw} & Basic water service
 & Defined as an improved water source with either water on the premises or round-trip collection time is 30 minutes or less.
\\\midrule
 \texttt{liws} & Limited water service
 & Limited water service, defined as an improved water source with round-trip collection time greater than 30 minutes.
\\\midrule
 \texttt{prem} & Water on the premises
 & 
\\\midrule
 \texttt{bicy} & Bicycle
 &
\\\midrule
 \texttt{acar} & Animal drawn cart
 & 
\\\midrule
 \texttt{mcyc} & Motorcycle
 & 
\\\midrule
 \texttt{pcar} & Private car
 & 
\\\midrule
 \texttt{boat} & Boat with a motor
 & 
\\\midrule
 \texttt{wigc} & Wealth index Gini coefficient
 & The Gini coefficient indicates the level of concentration of wealth in the country.
  \\\bottomrule
  \end{tabular}
  \caption{Summary of explanatory variables (National Household usage Averages measured as percentage for Main Drinking Source) }
  \label{tab:summary-var}
\end{table}
 \subsection*{Summary of variables}
\hl{Here, you show and discuss the data distributions (e.g. histograms, correlation plots, etc)}

@@

We analyzed data from the Demographic Health Survey (DHS) and the Malaria Indicator Survey (MIS) (1990-2020) to highlight water accessibility predictors in 78 countries throughout the world.

@@

\subsection{Exploratory Data Analysis}
First, we developed a correlation chart to quickly analyze if there is any great correlation between all the 28 numerical variables that may pertain to water accessibility. We recognized the strong and weak relationship by the size and color of the correlation plot is shown in \autoref{fig:cor-basic}, which indicates households with limited water service has the strongest and positive relationship with Households possessing a bicycle. Meanwhile, households using an improved water source has the strongest and negative relationship with households using an unimproved water source. This initial finding results indicated that the initial variables are logical and can give us insight into further understanding and the relationship between these variables.

@@

NBClust statistic analysis developed and determined four distinct groups. The first cluster group is composed of countries near the equator or in the Sub-Sahara Regions. The analysis of the second group represents countries considered developed or recently industrialized. The third cluster group represents most of the countries in central Africa and may have similar factors as the first cluster group. The fourth cluster group represents most of the countries in the middle east or predominately non-African countries. There are great internally heterogeneous clusters for the 76 distinct countries. The cluster analysis' dendrogram is shown in \autoref{fig:dendrogram}, and the corresponding NBClust curve yielding the optimal four clusters is shown in \autoref{fig:optimal-clusters}, \autoref{fig:optimal}. \autoref{tab:water-cluster} shows the resulting four water accessibility country clusters. The cluster's water accessibility variables are summarized in \autoref{tab:summary-wa}. \autoref{fig:cluster-map} shows the cluster country membership of the water accessibility on a world map.  Through the optimal analysis of the clusters, which indicated the superior grouping of four cuts along with the relevant indicators.
